---
title: "Introduction to Math for Political Scientists"
author: "J. Alexander Branham & Megan Moeller"
date: "Fall 2015"
header-includes: 
   - \AtBeginPart{}
   - \AtBeginSection{}
   - \AtBeginSubsection{}
   - \AtBeginSubsubsection{}
output: 
  beamer_presentation:
    colortheme: beaver
    fig_caption: yes
    incremental: yes
    slide_level: 3
    theme: Warsaw
---


# Arithmatic 
## Addition and Subtraction
### Let's start real slow...
* So we all know that
$$2+2 =4$$
* But addition and subtraction have some cool (grool?) rules

### Properties of Addition and subtraction
* Communiciative:
    + $a\pm b = b \pm a$
* Associative
    + $(a\pm b) \pm c = a \pm (b\pm c)$
    
## Multiplication and Division
### Multiplication
* Multiplication - I have these 4 things 10 times. 
```{r} 
4+4+4+4+4+4+4+4+4+4 
```
\pause 
Or I could just do 
```{r}
4*10
```

### Division 
* Just fancy multiplication.
\pause
* I have these four things one of ten times.
```{r}
4*(1/10)
```

## Properties 
### Properties of Multiplication and Division
* Communicative 
    + $a*b = b*a$
* Associative
    + $(ab)c = a(bc)$
* Distributive
    + $a(b+c) = ab + ac$
    + Note that this works for division: $\dfrac{a+b}{c} = \dfrac{a}{c} + \dfrac{b}{c}$


# Properties of Equalities
### Relationships that hold with (real) numbers
* $a=b \longleftrightarrow b=a$ (Symmetric relationships)
* $a = b$ and $b=c \Rightarrow a=c$ (Transitive relationships)
    + $a>b$ and $b>c \Rightarrow a>c$
    
# PEMDAS
### PEMDAS
* Parentheses 
* Exponents
* Multiplication and division (left tor right)
* Addition and subtraction (left to right)
$$(10-48\div 12 * 2)^2 + 3^2 * (8-6)$$



