---
title: "Introduction to Math for Political Scientists"
author: "J. Alexander Branham & Megan Moeller"
date: "Fall 2015"
header-includes: 
   - \AtBeginPart{}
   - \AtBeginSection{}
   - \AtBeginSubsection{}
   - \AtBeginSubsubsection{}
output: 
  beamer_presentation:
    theme: "Warsaw"
    colortheme: "beaver"
    slide_level: 3
    incremental: TRUE
---

# Arithmatic 
## Addition and Subtraction
### Let's start real slow...
* So we all know that
$$2+2 =4$$
* But addition and subtraction have some cool (grool?) rules

### Properties of Addition and subtraction
* Communiciative:
    + $a\pm b = b \pm a$
* Associative
    + $(a\pm b) \pm c = a \pm (b\pm c)$
    
## Multiplication and Division
### Multiplication
* Multiplication - I have these 4 things 10 times. 
```{r} 
4+4+4+4+4+4+4+4+4+4 
```
\pause 
Or I could just do 
```{r}
4*10
```

### Division 
* Just fancy multiplication.
\pause
* I have these four things one of ten times.
```{r}
4*(1/10)
```

## Properties 
### Properties of Multiplication and Division
* Communicative 
    + $a*b = b*a$
* Associative
    + $(ab)c = a(bc)$
* Distributive
    + $a(b+c) = ab + ac$
    + Note that this works for division: $\dfrac{a+b}{c} = \dfrac{a}{c} + \dfrac{b}{c}$


# Properties of Equalities
### Relationships that hold with (real) numbers
* $a=b \longleftrightarrow b=a$ (Symmetric relationships)
* $a = b$ and $b=c \Rightarrow a=c$ (Transitive relationships)
    + $a>b$ and $b>c \Rightarrow a>c$
    
# PEMDAS
### PEMDAS
* Parentheses 
* Exponents
* Multiplication and division (left tor right)
* Addition and subtraction (left to right)
$$(10-48\div 12 * 2)^2 + 3^2 * (8-6)$$

# Exponents and Logarithms 
## What are they?
### Exponents and Logarithms
* Exponents tell you to multiply that thing by its base x times:
    + $3^4 = 3*3*3*3 = $
* Logarithms ask how many times you must raise the base to get x:
    + $log_3(81) = 4$
    + Note that logarithms with negative arguments are undefined 
    + Sometimes $log(n)$ means $log_{10}(n)$
    + Othertimes, it means $log_e(n) = ln(n)$
    
## Properties
### Properties of exponents
* $a^ma^n = a^{m+n}$
* $\dfrac{a^m}{a^n}  = a^{m-n}$
* $(a^m)^n = a^{mn}$
* $a^0=1$
* $a^{1/n} = \sqrt[n]{a}$
* $\left( \dfrac{a}{b} \right)^n = \dfrac{a^n}{b^n} = a^nb^{-n} \qquad \forall b\neq 0$
* $\left( \dfrac{a}{b} \right)^{-n} = \left( \dfrac{b}{a} \right)^n \qquad \forall a,b\neq 0$

### Properties of logs
* $log_x(ab) = log_xa + log_xb$
* $log_x(\dfrac{a}{b}) = log_xa - log_xb$
* $log_xa^b = b log_x a$
* $log_x 1 = 0$
* $m^{log_m(a)} = a$
* $\dfrac{log_x n}{log_x m} = log_m n$