---
title: "Introduction to Probability"
author: "J. Alexander Branham"
date: "Fall 2015"
header-includes: 
   - \AtBeginPart{}
   - \AtBeginSection{}
   - \AtBeginSubsection{}
   - \AtBeginSubsubsection{}
output: 
  beamer_presentation:
    colortheme: beaver
    fig_caption: yes
    incremental: yes
    slide_level: 3
    theme: Warsaw
---

```{r, include=FALSE}
library(ggplot2)
```


# What is probability?
### Probability
* Frequency with which an event occurs
* Typically:
$$
Pr(A) = P(A) = \pi(A) = \dfrac{\text{Number of ways an event can occur}}{\text{Total number of possible outcomes}}
$$
* There are other ways of thinking about probability, but we'll stick with this one

### Kolmogorov's Axioms 
* First: $Pr(E)\in\mathbb{R}, Pr(E)\geq 0 \qquad \forall E\in F$
    + Probabilities must be non-negative
* Second: $Pr(\Omega) = 1$
    + Something has to happen
    + Probabilities sum/integrate to 1
* Third: $Pr\left(\bigcup_{i = 1}^\infty E_i\right) = \sum_{i=1}^\infty Pr(E_i)$
    + The probability of disjoint (mutually exclusive) sets is equal to their sums 

# Probability Distributions
## Discrete
### Discrete probabilities 
* What's the probability that we'll roll a 3 on one die roll:
$$
Pr(y=3) = \dfrac{1}{6}
$$

* If one roll of the die is an "experiment"
* We can think of a 3 as a "success"
* And $Y \sim Bern(\dfrac{1}{6})$
* Fair coins are $\sim Bern(.5)$ for example
* More generally $Bern(\pi )$
    + $\pi$ represents the probability of success

### Binomial distributions
* Before we ran the experiment just once
* What happens if we run it multiple times?
$$
Pr(y_1=3, y_2=3) = \dfrac{1}{6} * \dfrac{1}{6}
$$

* Now $Y \sim Binom(2, \dfrac{1}{6})$
    + Generally, $Y \sim Binom(n, p)$
    + n = number of trials, p = probability of success
* PMF:
$$
{n \choose k} p^k (1-6)^{n-k}
$$

## Continuous distributions
### The basics
* What happens when our outcome is continous
* Much harder to think about... 
* There are infinity possible outcomes
* Makes the denominator of our fraction difficult to work with 
* What's the probability that you make $15,293.19/year? 
* Probability of the whole space must equal 1
* Even if all events equally likely, $\dfrac{1}{\infty} =0$ 
    + Kinda...
    
### Continuous distributions 
```{r, fig.height=2.25, fig.width=3}
ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  stat_function(fun = dnorm)
```

