---
title: "Intro to R - the Language"
author: "J. Alexander Branham"
date: "Fall 2015"
header-includes: 
   - \AtBeginPart{}
   - \AtBeginSection{}
   - \AtBeginSubsection{}
   - \AtBeginSubsubsection{}
   - \beamertemplatenavigationsymbolsempty
output: 
  beamer_presentation:
    theme: "Warsaw"
    colortheme: "beaver"
    slide_level: 3
    incremental: true
---
### 
> If you wanted to do foundational research in statistics in the mid-twentieth century, you had to be bit of a mathematician, whether you wanted to or not. . . . if you want to do statistical re search at the turn of the twenty-first century,
you have to be a computer programmer. - Andrew Gelman

# What is R?
## Intro
### What is R? 
* R is a language and environment for statistical computing and graphics.
    + From R's website
* Note that it is a **language**
    + Those of you who have learned a foreign language know how slow and frustrating that process is... 

## Why use R?
### Advantages of R
* It's **FREE**, open-source, and available on every platform
    + Free as in speech and free as in beer
    + Stata is between $55/yr and $1000/year for students, depending on what you need 
* Unparalleled in the number of statistical packages (groups of functions)
* Great community help (e.g. stackoverflow.com, R-bloggers)
* You can combine documents and r code together (rmarkdown or knitr)
    + For example, this presentation is written in rmarkdown

### Disadvantages of R
* Some feel like it's not as intuitive as Stata or other stats programs
* Uses lots of memory 
* If you know another programming language (C, Python, etc), R's syntax is very odd
* Since anyone can write a package, some of the code (and help files) are difficult to follow 

## Installation 
### R
* Go to www.r-project.org, click "Download R," select a mirror close to you, and select your operating system 
* R by itself is pretty dinky, so we like to use RStudio

### RStudio
* Go to www.rstudio.com and download the appropriate desktop version of Rstudio (the free one)
* Install like normal 
* Open RStudio, click tools > global options and change "Save workspace to RData on exit" to Never

# RStudio Layout
### RStudio Layout 
* Bottom left: Console
    + This is where you can enter R code to execute
* Top left: editor window
    + You can have open R scripts or rmarkdown files here to edit
* Top right: environment, history, (git)
    + Environment will list everything that R is "remembering"
    + History lists all commands entered in that "project"
* Bottom right: files, plots, packages, help

### R Projects
* You can set up different "projects" from within Rstudio
* This will automatically change the working directory to where you put the project
* I use a project for each paper, for example

# Using R
## The basics 
### Calculator 
* R is a great calculator 
```{r}
3 + 2
1.7729^4 * (1930/4)
```

### Assignment
R uses ` <- ` for assignment. 
```{r}
a <- 4
a + 7
a <- a + 2
a + 7
```
`a` is now referred to as an "object." Pretty much anything R remembers is an object. 

### Functions
Functions take arguments
```{r}
myvector <- c(1, 5, 2, 7, 9, NA, 1)
mean(myvector, na.rm = TRUE)
rnorm(5, 0, 1)
```

## Data Structures
### Basic Data Structures
R has four basic types of data: logical, numeric, integer, and character
```{r, eval = FALSE}
TRUE ; FALSE
3
3L
"character"
```


### matrix
```{r}
mymatrix <- matrix(c(1,2,3, 5,11,4), nrow=2, byrow=FALSE)
mymatrix
mean(mymatrix[ , 1])
```

### data.frame
```{r}
mydata <- data.frame(x=c(1,2,3), y=c(5,11,4))
mydata
mean(mydata$x)
```

### list
```{r}
mylist <- list(amatrix = mymatrix,
               adataframe = mydata)
mylist
```


## Programming tools
### If-else
If else statements say IF this is true, then do this. OTHERWISE, do that. There are two types in R: if, else (which work for scalars) and ifelse, which works on vectors
```{r}
x <- 1:10
if(5>3){
  x+1
}
ifelse(x<=5, x+1, x-4)
```

### For loops
For loops do something a specified number of times:
```{r}
x <- numeric()
for(i in 1:10){
  x[i] <- rnorm(1)
}
x
```

# Further Resources
### Further resources 
* [Importing data into R](http://blog.datacamp.com/r-data-import-tutorial/)
* [Useful R packages](https://support.rstudio.com/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages)
* R tutorials on lynda.com (free for UT students)
