---
title: "Calculus"
author: "J. Alexander Branham"
date: "Fall 2016"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation:
    fig_caption: yes
    incremental: yes
    slide_level: 3
    latex_engine: xelatex
classoption: aspectratio=169
fontsize: bigger
---

```{r, echo=FALSE}
library(ggplot2)
```
# Limits

### Limits
The limit of $f(x)$ as $x \rightarrow a$ is the value that $f(x)$
approaches as $x$ gets arbitrarily close to $a$

$$\lim_{x \to a}f(x) = L$$

### Limit example

Consider the following function:

$$f(x) = \frac{x^2-1}{x-1}$$

What is $f(1)$? 

### Limit example

$f(1)$ is undefined

```{r, fig.height=4}
ggplot(data.frame(x = c(0, 2)), aes(x)) +
  stat_function(fun = function(x) (x ^ 2 - 1) / (x - 1))
```

### Limits and infinity

$$\lim_{x \to \infty}x^2$$

## Limits and infinity 

```{r, fig.height=4}
ggplot(data.frame(x = c(0, 100000)), aes(x)) +
  stat_function(fun = function(x) x ^ 2)
```


### You try! 

$$\lim_{x \to -\infty}\frac{1}{x}$$

### You try! 

```{r, fig.height=4}
ggplot(data.frame(x = c(-10000, -1)), aes(x)) +
  stat_function(fun = function(x) ifelse(x != 0, 1 / x, NA)) 
```

### Sided limits

What is 

$$\lim_{x \to 0} \frac{1}{x}$$

### Sided limits

```{r, fig.height=4}
ggplot(data.frame(x = c(-1, 1)), aes(x)) +
  stat_function(fun = function(x) ifelse(x != 0, 1 / x, NA)) 
```

### Sided limits

Here, we can take two different limits: the limit as $x$ approaches
$0$ from the left and then from the right

$$\lim{x\to 0^{-}} \qquad \text{from the left}$$
$$\lim{x\to 0^{+}} \qquad \text{from the right}$$


### Sided limits
The limit does not exist! 

\pause

For a limit to exist, both sides must exist *and* they must be equal

### You try! 

Find both one sided limits and the two sided limit as $x \to 1$ of $f(x)$

$$f(x) = \begin{cases}
x^2 & \text{if } x < 1 \\
x-3 & \text{if } x \geq 1\\
\end{cases}$$


### Properties of limits

$$\lim_{x\to a} c = c$$

\pause

$$\lim_{x\to a} x=a$$

\pause

$$\lim_{x\to a}\left(f(x) \pm g(x) \right) = \lim_{x\to a} f(x) \pm
\lim_{x\to a} g(x)$$

### Properties of limits, continued

$$\lim_{x\to a}\left( \frac{\lim_{x\to a} f(x)}{g(x)} \right) =
\frac{\lim_{x\to a} f(x)}{\lim_{x\to a} g(x)} \qquad \text{iff }
\lim_{x\to a}g(x) \neq 0 $$

\pause

$$\lim_{x\to a}\left( c f(x)\right) = c \lim_{x\to a}$$

\pause

$$\lim_{x\to a}\left( f(x)^n \right) = \left(\lim_{x\to
a}f(x)\right)^n \qquad \text{iff } \lim_{x\to a}f(x) > 0$$

# Derivatives 
## Derivatives 

### Derivatives
Derivatives calculate the instantaneous slope (rate of change) of a
function at every point on its domain

$$\frac{dy}{dx} = \frac{d}{dx}f(x) = f'(x) = \lim_{h\to 0 }
\frac{f(x+h)-f(x)}{h}$$

### The one derivative rule to rule them all 

$$\frac{dy}{dx} x^n = nx^{n-1}$$

## Optimization

## Partial derivatives

# Integrals 
