---
title: "Calculus"
author: "J. Alexander Branham"
date: "Fall 2016"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: beamer_presentation
classoption: aspectratio=169
fontsize: bigger
---
```{r, echo=FALSE}
source("r-setup.R")
```
# Limits

### Limits
The limit of $f(x)$ as $x \rightarrow a$ is the value that $f(x)$
approaches as $x$ gets arbitrarily close to $a$

$$\lim_{x \to a}f(x) = L$$

### Limits
Let $f(x)$ be $x+3$. Find:

$$\lim_{x\to 3}f(x)$$

### Limits

```{r}
ggplot(data.frame(x = c(-1, 5)), aes(x)) +
  stat_function(fun = function(x) x + 3) +
  geom_vline(xintercept = 3, linetype = "dashed")
```

### Limit example

Consider the following function:

$$f(x) = \frac{x^2-1}{x-1}$$

What is $f(1)$? 

### Limit example

$f(1)$ is undefined

```{r}
ggplot(data.frame(x = c(0, 2)), aes(x)) +
  stat_function(fun = function(x) (x ^ 2 - 1) / (x - 1))
```

### Limits and infinity

$$\lim_{x \to \infty}x^2$$

## Limits and infinity 

```{r}
ggplot(data.frame(x = c(0, 100000)), aes(x)) +
  stat_function(fun = function(x) x ^ 2)
```


### You try! 

$$\lim_{x \to -\infty}\frac{1}{x}$$

### You try! 

```{r}
ggplot(data.frame(x = c(-10000, -1)), aes(x)) +
  stat_function(fun = function(x) ifelse(x != 0, 1 / x, NaN)) 
```

### Sided limits

What is 

$$\lim_{x \to 0} \frac{1}{x}$$

### Sided limits

```{r}
ggplot(data.frame(x = c(-1, 1)), aes(x)) +
  stat_function(fun = function(x) ifelse(x != 0, 1 / x, NaN)) 
```

### Sided limits

Here, we can take two different limits: the limit as $x$ approaches
$0$ from the left and then from the right

$$\lim_{x\to 0^{-}} \qquad \text{from the left}$$
$$\lim_{x\to 0^{+}} \qquad \text{from the right}$$


### Sided limits
The limit does not exist! 

\pause

For a limit to exist, both sides must exist *and* they must be equal

### You try! 

Find both one sided limits and the two sided limit as $x \to 1$ of $f(x)$

$$f(x) = \begin{cases}
x^2 & \text{if } x < 1 \\
x-3 & \text{if } x \geq 1\\
\end{cases}$$

### You try (answers)

```{r, warning=FALSE}
ggplot(data.frame(x = -3:5), aes(x)) +
  stat_function(fun = function(x) ifelse(x < 1, x ^ 2, NA)) +
  stat_function(fun = function(x) ifelse(x >= 1, x - 3, NA)) +
  geom_vline(xintercept = 1, linetype = "dashed")
```

### Properties of limits

$$\lim_{x\to a} c = c$$

\pause

$$\lim_{x\to a} x=a$$

\pause

$$\lim_{x\to a}\left(f(x) \pm g(x) \right) = \lim_{x\to a} f(x) \pm
\lim_{x\to a} g(x)$$

### Properties of limits, continued

$$\lim_{x\to a}\left( \frac{\lim_{x\to a} f(x)}{g(x)} \right) =
\frac{\lim_{x\to a} f(x)}{\lim_{x\to a} g(x)} \qquad \text{iff }
\lim_{x\to a}g(x) \neq 0 $$

\pause

$$\lim_{x\to a}\left( c f(x)\right) = c \lim_{x\to a} f(x)$$

\pause

$$\lim_{x\to a}\left( f(x)^n \right) = \left(\lim_{x\to
a}f(x)\right)^n \qquad \text{iff } \lim_{x\to a}f(x) > 0$$

# Derivatives 
## Derivatives 

### Derivatives
Derivatives calculate the instantaneous slope (rate of change) of a
function at every point on its domain

Can think of this like so:

$$\lim_{\Delta x \to 0} \frac{\Delta y}{\Delta x}$$

### Derivatives - notation 

$$\frac{dy}{dx} = \frac{d}{dx}f(x) = f'(x) = \lim_{h\to 0 }
\frac{f(x+h)-f(x)}{h}$$

### Other derivatives 

$$\frac{dy}{dx}c = 0$$
\pause
$$\frac{d}{dx}e^x = e^x$$
\pause
$$\frac{d}{dx}\log_e(x) = \frac{1}{x}$$


### The one derivative rule to rule them all 
The **power rule**

$$\frac{dy}{dx} x^n = nx^{n-1}$$

### Other derivative rules
$$\frac{d}{dx}cf(x) = cf'(x)$$
\pause 
$$(f(x) \pm g(x))' = f'(x) \pm g'(x)$$
\pause 
$$(f(x)g(x)) = f(x)g'(x) + f'(x)g(x)$$

### The chain rule
$$\frac{d}{dx}f(g(x)) = f'(g(x))g'(x)$$

### Higher-order derivatives
The "second derivative" is just the derivative of the derivative:

$$f''(x) = \frac{d^2}{dx}$$

You can take the third derivative and so on as well

### You try!

Find the following:

1. $\frac{d}{dx} x^2$
2. $\frac{d}{dx} x^3 + x^2 + 2x - 1$
3. $\frac{d^2}{dx} x^4 + x^3$

## Optimization
### Optimization
Sometimes we want to find the maximum (or minimum) value a function
takes. **Optimization** lets us do this!

### Optimization
The derivative of a function gives rate of change, so when that is
zero, the function has usually reached a (local) maximum or minimum.
Why? 

So we can find the local maxima and/or minima of a function by taking the
derivative, setting it equal to zero, and solving for $x$ (or
whatever)

### Optimization
**BUT** we don't know if we've found a maximum or minimum. 

The second derivative gives us the rate of change *of the rate of
change* of the original function. So it tells us whether the slope is
getting larger or smaller. 

$$f(x) = x^2$$
$$f'(x) = 2x$$
$$f''(x) = 2$$

So a *positive* second derivative indicates a *minimum* and a
*negative* second derivative indicates a *maximum*

### Local vs global optimization
To find the max/min on some interval you have to compare the local
max/min to the value of the function at the interval's endpoints. To
find global max/min you must check the function's limits as it
approaches positive and negative $\infty$

### You try!
Find the global minimum for the following function:
$$f(x) = x^2 - 2x + 7$$

### You try (answer)
Find first derivative:
$$f'(x) = 2x - 2$$
\pause
Set it equal to zero
$$2x - 2 = 0$$
$$x = 1$$
\pause 
Check the second derivative to see if this is a maximum or minimum:
$$f''(x) = 2$$
It's positive, so this is a *minimum*.
\pause
Check the "endpoints"
$$\lim_{x\to -\infty}f(x) = \infty$$
$$\lim_{x\to \infty}f(x) = \infty$$


### You try (answer)
```{r}
ggplot(data.frame(x = c(-3, 3)), aes(x)) +
  stat_function(fun = function(x) x ^ 2 - (2 * x) + 7) +
  geom_vline(xintercept = 1, linetype = "dashed")
```

## Partial derivatives
### Partial derivatives
Some functions have more than one variable (e.g. $f(x,z) = x^2 +2xz +
z^2$). In this case, we have to be very clear with respect to which we
are differentiating. We can only do one at a time and when we do so w
treat all the other variables as constants. The notation is:

$$\frac{\partial f(x,z)}{\partial x}$$

### Partial derivatives 
Example: $y = 3x^2w-4w$
\pause
$$\frac{\partial y}{\partial x} = 6xw$$
\pause 
$$\frac{\partial y}{\partial w} = -4$$


# Integrals 
## Integrals
### Geometry

Sometimes (pretty often, turns out) we might be interested in
obtaining the **area under a curve**

This is sometimes super easy. What is the area under the curve between
$x = -1$ and $x = 1$? 

$$
f(x) = \begin{cases}
    \frac{1}{3} & \text{for } x \in [0, 3] \\
    0 & \text{otherwise} \end{cases}
$$

### Geometry 

$$A = \ell w$$

```{r, echo = FALSE, warning = FALSE}
ggplot(data.frame(x = -1:4), aes(x)) +
  stat_function(fun = function(x) ifelse(x > 0 & x < 3, (1 / 3), NaN)) +
  stat_function(fun = function(x) ifelse(x < 0 | x > 3, 0, NaN)) +
  geom_rect(aes(xmin = 0, xmax = 1,
                ymin = 0, ymax = (1 / 3)), fill = "grey", alpha = 0.3) +
  coord_cartesian(ylim = c(-1, 1))
```

### Integration

This is the basic question motivating integration. If we wanted to
find the area under the curve of $f(x)$ from $x=0$ to $x=4.5$, this is
the notation:

$$\int_{x=0}^{4.5} f(x) \, dx$$

\pause

How do we do this? 

### Finding integrals 

Consider: $f(x) = -x^2 + 7$

```{r}
ggplot(data.frame(x = -5:5), aes(x)) +
  stat_function(fun = function(x) -x ^ 2 + 7)
```

How to find the area between 0 and 2?

### Riemann Integrals
\centering
((Board examples))

### Integration rules
Where $a$ is a constant:

$$\int a \, dx = ax + C$$

\pause 

$$\int af(x) \, dx = a \int f(x) \, dx$$

\pause 

$$\int (f(x) + g(x)) \, dx = \int f(x) \, dx + \int g(x) \, dx$$

### Integration rules

$$\int x^n \, dx = \frac{x^{n+1}}{n+1} + C \qquad \forall n \neq 0$$


$$\int x^{-1} \, dx = \ln |x| + C$$

### You try!
Find the integral:

1. $\int x^2 \, dx$
2. $\int 3x^2 \, dx$
3. $\int 3x^2 + 2x - 7 \, dx$
