---
title: " Matrix Algebra"
author: "J. Alexander Branham"
date: "Fall 2017"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: beamer_presentation
classoption: aspectratio=169
fontsize: bigger
---
```{r, echo=FALSE}
source("r-setup.R")
```

# What is a matrix?

## Scalars

### Scalars
* Let's start with something familiar, with a new word
* One number (12, for example) is referred to as a *scalar*
* This can be thought of as a $1 x 1$ matrix
    + More on that in a bit...
$$
\begin{bmatrix}
12
\end{bmatrix}
= c  $$

## Vectors 

### Vectors 
* We can put several scalars together to make a vector
* For example: 
$$
\begin{bmatrix}
12 \\
14 \\
15
\end{bmatrix}
= b
$$
* Since this is a column of numbers, we cleverly refer to it as a *column vector*

### Row Vectors 
If we take $b$ and arrange it so that it it a row of numbers instead of a column, we refer to it as a *row vector*:
$$
\begin{bmatrix}
12 & 14 & 15
\end{bmatrix}
= d
$$

# Operators
### Operations on vectors:

The **summation operator** $\sum$ lets us perform an operation (sum)
on a sequence of numbers (often but not always a vector):

$$\sum_{i=1}^{5} i$$


### Summation operator
Let:

$$x_i = \begin{bmatrix}
12 & 7 & -2 & 0 & 1
\end{bmatrix}$$

Find: 

$$
\sum_{i=1}^3 x_i
$$

### Summation operator

```{r}
x <- c(12, 7, -2, 0, 1)
sum(x[1:3])
```
### You try! 
Let 
$$ y = \begin{bmatrix} 
1 & 0 & -1 & 4
\end{bmatrix} $$

Find: 

$$\sum_{i = 1}^\infty y^2$$

### You try (answer)
```{r}
y <- c(1, 0, -1, 4)
sum(y ^ 2)
```

### Product operator
We might want to multiply instead of add, in which case we can use the
product operator $\prod$

### Product operator
Let: 
$$z = \begin{bmatrix}
6 & -2 & 0 & 1
\end{bmatrix}$$

Find: 

$$
\prod_{i=1}^{2} z_i
$$

### Product operator

```{r}
z <- c(6, -2, 0, 1)
prod(z[1:2])
```

### You try!
Let:

$$a = \begin{bmatrix}
1 & 2 & 3 & 4 & 5 & 4 & 3 & 2 & 1
\end{bmatrix}$$

Find:

$$\prod_{i=1}^{3} a^2 -2a + 3$$

### You try (answer)

```{r}
a <- c(1, 2, 3, 4, 5, 4, 3, 2, 1)
prod(a[1:3] ^ 2 - 2 * a[1:3] + 3)
```

# Matrices 
## Matrices
### Matrix
We can put multiple vectors together to get a *matrix*:
$$
\begin{bmatrix}
12 & 14 & 15 \\
115 & 22 & 127 \\
193 & 29 & 219
\end{bmatrix}
= A
$$

### Matrices, cntd
* We refer to the *dimensions* of matrices by *row* x *column*
* So $A$ is a $3x3$ matrix.
* Note that matrices are usually designated by capital letters 
    + And sometimes bolded as well
    
### Dimensions
\centering
\Huge ROW x COLUMN

### Indices 
* How do we refer to specific elements of the matrix???
* Solution: come up with a clever indexing scheme
* Matrix $A$ is an $nxm$ matrix where $n=m=3$.
* More generally, matrix $B$ is an $n\times m$ matrix where the elements look like this:
$$
B=
\begin{bmatrix}
b_{11} & b_{12} & b_{13} & \ldots & b_{1m} \\
b_{21} & b_{22} & b_{23} & \ldots & b_{2m} \\
\vdots & \vdots & \vdots & \ldots & \vdots \\
b_{n1} & b_{n2} & b_{n3} & \ldots & b_{nm}
\end{bmatrix}
$$

# Matrix Operations
## Addition and subtraction

### Addition and subtraction are EASY!
* Requirement: Must have \emph{exactly} the same dimensions
* To do the operation, just add or subtract each element with the corresponding element from the other matrix:
$$
A \pm B
$$

### Addition and Subtraction
$$
\begin{bmatrix}
a_{11} & a_{12} & a_{13}\\
a_{21} & a_{22} & a_{23}\\
a_{31} & a_{32} & a_{33}
\end{bmatrix}
\pm
\begin{bmatrix}
b_{11} & b_{12} & b_{13}\\
b_{21} & b_{22} & b_{23}\\
b_{31} & b_{32} & b_{33}
\end{bmatrix}
$$

$$
=
\begin{bmatrix}
a_{11}\pm b_{11} & a_{12}\pm b_{12} & a_{13}\pm b_{13}\\
a_{21}\pm b_{21} & a_{22}\pm b_{22} & a_{23}\pm b_{23}\\
a_{31}\pm b_{31} & a_{32}\pm b_{32} & a_{33}\pm b_{33}
\end{bmatrix}
$$

### You try! 

Let

$$ A= \begin{bmatrix}
1 & 4 & 2 \\
-2 & -1 & 0 \\
0 & -1 & 3
\end{bmatrix}
B = \begin{bmatrix}
5 & 1 & 0 \\
2 & -1 & 0 \\
7 & 1 & 2
\end{bmatrix}$$

Find $A + B$

### You try (answers)

```{r}
A <- matrix(c(1, 4, 2, -2, -1, 0, 0, -1, 3),
            byrow = TRUE, nrow = 3)
B <- matrix(c(5, 1, 0, 2, -1, 0, 7, 1, 2),
            byrow = TRUE, nrow = 3)
A + B
```

## Multiplication

### Scalar multiplication
Easy - just multiply each element of the matrix by the scalar
$$
cA = 
\begin{bmatrix}
ca_{11} & ca_{12} & ca_{13}\\
ca_{21} & ca_{22} & ca_{23}\\
ca_{31} & ca_{32} & ca_{33}
\end{bmatrix}
$$

### Scalar multiplication
```{r}
A
3 * A
```

### Matrix multiplication
* Requirement: the two matrices must be *conformable*
* This means that the number of columns in the first matrix equals the number of rows in the second
* The resulting matrix will have the number of rows in the first, and the number of columns in the second! 

### Pop quiz
* Which can we multiply? What will the resulting dimensions be?
$$
b=
\begin{bmatrix}
2 \\
3\\
4\\
1
\end{bmatrix}
M = 
\begin{bmatrix}
1 & 0 & 2\\
1 & 2 & 4\\
2 & 3 & 2
\end{bmatrix}
L = 
\begin{bmatrix}
6 & 5 & -1\\
1 & 4 & 3 
\end{bmatrix}
$$

* **ONLY** $LM$ and **NOT** $ML$
* The dimensions will be $2x3$

### How to actually do this? 
* Multiply each row by each column
* (Board examples)

\pause 

```{r}
L <- matrix(c(6, 5, -1, 1, 4, 3),
            nrow = 2, byrow = TRUE)
M <- matrix(c(1, 0, 2, 1, 2, 4, 2, 3, 2),
            nrow = 3, byrow = TRUE)
L %*% M
```



## Division
### Matrix Division
\centering 
\pause
\Huge HAHAHA... NOPE

## Properties of matrix operations
### Properties of operators 
* Addition and subtraction
    + Associative $(A \pm B) \pm C = A \pm (B \pm C)$
    + Communicative $A \pm B = B \pm A$
* Multiplication
    + $AB \neq BA$
    + $A(BC) = (AB)C$
    + $A(B+C) = AB + AC$
    + $(A+B)C = AC + BC$

# Transposition
## What is it? 
### Qu'est-ce que c'est?

* Switch the rows and columns
* So a $nxm$ matrix becomes $mxn$
* Typically denoted $L'$ or $L^T$

### Transposition 

```{r}
L
t(L)
```

## Properties 
### Properties of transposing
* Matrix is *always* conformable for multiplication with its transpose
  in both directions
* $(A\pm B)' = A' \pm B'$
* $A'' = A$
* $(AB)' = B'A'$
* $(cA)' = cA'$ where $c$ is a scalar
* $AA'$ and $A'A$ will always result in a symmetric matrix

### You try!
Let: 

$$
A = \begin{bmatrix}
2 & 4 \\
3 & 1
\end{bmatrix}
B = \begin{bmatrix}
2 & 3 & 4 \\
-1 & 0 & 1
\end{bmatrix}
$$

Find: 

1. $A'A$
2. $AB$
3. $BA$

### You try (answers)

```{r} 
A <- matrix(c(2, 4, 3, 1), byrow = TRUE, nrow = 2)
B <- matrix(c(2, 3, 4, -1, 0, 1), byrow = TRUE, nrow = 2)
A %*% t(A)
```

### You try (answers)

```{r}
A %*% B
```

### You try (answers)

```{r, error = TRUE}
B %*% A
```

# Matrix Inverse
## What is it? 
### Matrix Inverses
* I kinda lied when I said there isn't matrix division
* We use matrix inverses all the time
* If $A$ is an $nxn$ square matrix: 
$$
AB = BA = I_n
$$
* Then $B$ is said to be the *inverse* of $A$
    + This is usually denoted $A^{-1}$
    + So $AA^{-1} = I_n$
* If $B$ doesn't exist, then the matrix is *singluar*
* Finding inverses by hand is super hard (especially as $n$ increases), so we let computers do this for us 

## Properties of inverses 
### Some properties
* Let $A$ be $nxn$ square matrix:
* If $A^{-1}$ exists:
* $A$ is full rank: $rank(A) = n$
* $A'$ is also invertible 
* $(A^{-1})^{-1} = A$
* $(cA)^{-1} = c^{-1}A^{-1}$ for nonzero scalar $c$
* $(A')^{-1} = (A^{-1})'$

# Special matrices
## Special matrices
### Special types of matrices
Some matricies get more love than others

### Square matrix
Any $nxn$ matrix (same number rows and columns)
$$
\begin{bmatrix}
1 & 0 & 2\\
1 & 2 & 4\\
2 & 3 & 2
\end{bmatrix}
$$


### Symmetric matrix
A square matrix that is the same as its transpose
$$
\begin{bmatrix}
2 & 5 & 7\\
5 & 9 & 6\\
7 & 6 & 7
\end{bmatrix}
$$

### Diagonal matrix
A symmetric matrix with zeros everywhere but the main diagonal 
$$
\begin{bmatrix}
2 & 0 & 0\\
0 & 9 & 0\\
0 & 0 & 7
\end{bmatrix}
$$


### Scalar matrix 
A diagonal matrix with the same number all along the diagonal
$$
\begin{bmatrix}
2 & 0 & 0\\
0 & 2 & 0\\
0 & 0 & 2
\end{bmatrix}
$$

### Identity matrix
A scalar matrix where the diagonal elements are 1. 
$$
\begin{bmatrix}
1 & 0 & 0\\
0 & 1 & 0\\
0 & 0 & 1
\end{bmatrix}
$$
\pause

* This is a super important type of matrix. 

* It gets its own notation: $I_n$ where $n$ is the number of rows and columns

* Note that $I_n A = A$ and also $A I_n = A$

