---
title: "Set Theory & Combinations"
author: "J. Alexander Branham"
date: "Fall 2016"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation:
    fig_caption: yes
    incremental: yes
    slide_level: 3
    latex_engine: xelatex
classoption: aspectratio=169
fontsize: bigger
---
```{r, echo=FALSE}
source("r-setup.R")
```

# Set Theory
## What is it?
### Intro to Set Theory
* What is set theory?
* A branch of mathematics 
* Collects objects into *sets* and studies the properties 

### What's a set? 
* A **set** is a collection of objects 
$$
S = \left\{ s_1, s_2, s_3, ... s_n \right\}
$$
* The objects can be anything
* We usually use variables or units of observation 

### Elements in or not
* We can say whether an object is in a set:
$$
s_{13} \in S
$$
* Or not:
$$
q_{1} \notin S
$$

## Subsets
### Subsets
* We can divide sets into *subsets*
$$
M \subset S
$$
* $M$ is a *proper subset* of $S$ iff all elements of $M$ are in $S$ but not all elements of $S$ are in $M$. 
* We can also define non-proper subsets:
$$
L \subseteq S
$$

## Set universes
### Empty sets
* Some sets contain no elements
* For example, cities on Earth with average temperatures of greater than 1000 degrees
  + Though Austin might be close... 
$$
Z = \left\{ \emptyset \right\}
$$

### Set Universes
* We can talk about how subsets fill a certain **universe** of possibilities
* For example:
$$
R = \left\{1,2,3,4,5,6\right\}
$$
* $R$ represents all the possibilities of a (single) roll of a die 
* We can define sets for the even possibilities and the odd possibilities
$$
E = \left\{ 2,4,6 \right\} \qquad O = \left\{ 1,3,5 \right\}
$$

### Compliments
* A *compliment* is that together, they contain all the elements of the relevant universe 
$$
E = O^C \qquad \qquad O = E^C
$$

## Graphical representation
### Universe
* Board examples of how to draw sets

## Intersections and Unions
### Intersection, Disjoint, unions
* Intersections: $A \cap B$ includes all the common elements of $A$ and $B$
    + If $A \cap B = \emptyset$, then they are **disjoint**
* Unions: $A \cup B$ contains all the elements of both $A$ and $B$
    + Two sets unions can't be empty (unless both sets are empty to begin with)
* Size: $n(R)=6$ tells us that there are 6 elements in $R$
* Sizes of unions
    + $n(X \cup Y) = n(X) + n(Y) - n(X \cap Y)$
    + Have to subtract the last term, otherwise we'd double-count elements in the intersection of the two sets
    
### Properties
* $A \cup B = B \cup A$
* $A \cap B = B \cap A$
* $(A \cup B) \cup C = A \cup (B \cup C)$
* $(A \cap B) \cap C = A \cap (B \cap C)$
* $A \cap (B \cup C) = (A \cap B) \cup (A \cap C)$
* $A \cup (B \cap C) = (A \cup B) \cap (A \cup C)$

### You try!
1. Define the universe of cards in a standard 52-card deck --- $\Omega$
2. Define the set containing all the spades --- $S$ --- the set
   containing the black cards --- $B$ --- and the set of queens $Q$
3. What is $n(B)$
4. What is $S \cup B$? 
5. What is $B \cap Q$? 
6. What is $B \cap (Q \cup S)$? 
7. Is $Q \subset S$?

### You try (answers)
1. $\Omega$ is a set with 52 elements, one for each card
2. $S$ is a set with 13 elements, one for each spade. $B$ has 26
   elements (the black cards) and $Q$ has just four - the queens
3. $n(B) = 26$
4. $S \cup B = B$ 
5. $B \cap Q$ is the queen of spades and clubs
6. $B \cap (Q \cup S) = B$
7. No

# Combinatorics 
## What is it?
### Combinatorics - What is it? 
* How to count things? 
* Countable *discrete* structures 
* Let's say we have the set $R$ from above
$$
R = \left\{ 1,2,3,4,5,6 \right\}
$$
* How many different combinations of 3 dice rolls are there? 

### Order
* First thing to ask is if order matters
* Is rolling 1, 4, 2 the same as rolling 4, 2, 1?
* If order matters, we talk about **permutations**
* If not, it's a **combination**
* think "permutation" = "position"

## Permutations
### Repitition? 
* Can we repeat things? 
* Is rolling 3, 3, 3, permissable? 
* If yes, we choose one die roll of six three times:
$$
6 * 6 * 6 = 6^3
$$
* more generally, $n^r$
    + $n$ is number of things to choose from
    + choose $r$
    
### Permutations without replacement
* We need to reduce the number of things from which to choose each time
* How many different orderings of 9 horses in a race??
    + $9*8*7$...$=9!=362,880$
* Alternatively, maybe we just care about the first three:
    + $9*8*7=504$
* More generally:
$$
\dfrac{n!}{(n-r)!}
$$

## Combinations
### Combinations
* We use combinations when the order does **not** matter
* So 4, 3, 1 is the same as 3, 1, 4 

### Repition not allowed
* Easiest way: do permutation without repitition, then get rid of all the "same" outcomes
* We've already figured out the permutation part, so need to figure out the second part 
$$
\dfrac{n!}{(n-r)!} * \dfrac{1}{r!} = \dfrac{n!}{r!(n-r)!} = {n \choose r}
$$

### Repitition allowed
* These aren't used much
* Also the most difficult to calculate
* Formula:
$$
{n+r-1 \choose r} = \dfrac{(n+r-1)!}{r!(n-1)!}
$$

### You try!
Using a standard 52-card deck:

1. How many different 3-card hands exist? 
2. How many ways are there to get dealt the queen of spades, then the
   king of diamonds, then the eight of spades? 

### You try (answers)

```{r}
# 1
choose(52, 3)

# 2 - one
```
