---
title: "Statistics"
author: "J. Alexander Branham"
date: "Fall 2016"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation:
    fig_caption: yes
    incremental: yes
    slide_level: 3
    latex_engine: xelatex
classoption: aspectratio=169
---

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
```

# Statistics
### Statistics
*Statistics* allow us to learn from data

### Data
Observations of a *variable*: 

$$x = \begin{bmatrix}
1 \\ 0 \\ 0 \\ 1 \\ 0 \\ 0\\ 0 \\ 0 \\ 1
\end{bmatrix}$$

## Univariate statistics
### Univariate statistics
A **statistic** summarizes data. You're already familiar with some
common statistics, like averages. 

We oftentimes want to find the "center" of the data --- this describes
typical values

### Central tendency
$$x = \begin{bmatrix} 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 \end{bmatrix}$$
The *mean* ($\bar{x}$) is calculated by summing the data, then dividing by the
number of observations:
$$\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i$$
\pause
The *median* is found by ordering the observations from highest to
lowest and finding the one in the middle:
\pause 
The *mode* is the most common number

### Variance
Finding central tendency is good, but we might go a step further.
Consider these two distributions:

```{r, echo = FALSE, fig.height = 4}
ggplot(data.frame(x = -6:6), aes(x)) +
  stat_function(fun = dnorm) +
  stat_function(fun = dnorm, args = list(sd = 3), color = "red")
```

### Variance 
*Variance* measures how spread out a distribution is. One way to
calculate it is like so:

$$s^2_x = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2$$

That measures the sum of the squared average deviation from the mean 

### Standard deviation
"Squared average deviation from the mean" is a bit weird, though, so
oftentimes we use standard deviations instead, which is just the
squared root of the variance:

$$s_x = \sqrt{s^2_x}$$


