---
title: "Data visualization with R"
author: "J. Alexander Branham"
date: "Fall 2017"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation: 
      fig_width: 6
      fig_height: 6
classoption: aspectratio=169
fontsize: bigger
---

```{r setup, include=FALSE}
source("r-setup.R")
library(gapminder)
```

# Data visualization 
## Data visualization
### Data visualization 
* Plots are (usually) easier to interpret than tables and numbers
* We'll use `ggplot2` to visualize data in R. Make sure you have it
  installed (only the first time) and loaded (every time):

\pause

```{r, eval = FALSE}
install.packages(c("ggplot2", "gapminder"))
library(ggplot2, gapminder)
```

### Data visualization 
We'll be working with the `gapminder` dataset, provided by `gapminder`.
Check out its documentation with `?gapminder`:

1. How many rows (observations) are in the data?
2. What's the name of the variable for life expectancy? 
3. What's the name of the variable for GDP per capita? 


### Weirdness
Kuwait has a super high GDP per cap in some years. We're just going to
drop them for the purpose of this tutorial: 

```{r}
gapminder <- gapminder %>%
  filter(gdpPercap < 50000)
```


### Scatterplots

```{r, fig.show='hide'}
p <- ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp))
```

### Scatterplots

```{r, echo=FALSE}                      
p
```

# Aesthetics
## Aesthetics
### Color

```{r}
p <- ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap,
                           y = lifeExp,
                           color = continent))
```                                                                                   

### Color
```{r, echo=FALSE}
p
```

### Shape

```{r}
p <- ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap,
                           y = lifeExp,
                           shape = continent))
```

### Shape

```{r, echo=FALSE}
p
```

### Size

```{r, warning = FALSE}
p <- ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap,
                           y = lifeExp,
                           size = continent))
```

### Size

```{r, echo=FALSE}
p
```



### Alpha

```{r}
p <- ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap,
                           y = lifeExp,
                           alpha= continent))
```

### Alpha
```{r, echo=FALSE}
p
````

### You try!
Use the `mpg` dataset from `ggplot2` to:

1. Make a scatterplot with displ on the x-axis and hwy on the y-axis
2. Add color, size, and shape aesthetics

# Facets
### Facets
Can make facets by adding `facet_grid`:
                            
```{r}
p <- ggplot(data = gapminder) +
  geom_point(mapping = aes(x = gdpPercap, y = lifeExp)) +
  facet_grid(. ~ continent)
```

### Facets

```{r, echo=FALSE}
p
```

# Geoms
### Geoms
There are other kinds of graphs than scatterplots. `geom_` takes care
of this for ggplot2 ("geometric object"):

### Smooth
```{r}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_smooth()
```

### Smooth
```{r, echo=FALSE}
p
```


### Smooth - SE's

```{r}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent)) +
  geom_smooth(se = FALSE)
```

### Smooth - SEs
```{r, echo=FALSE}
p
```

### Smooth - OLS

```{r}
p <- ggplot(data = gapminder,
            mapping = aes(x = gdpPercap, y = lifeExp,
                          color = continent)) +
  geom_smooth(se = FALSE, method = "lm")
```

### Smooth - OLS
```{r, echo=FALSE}
p
```

### Histogram

```{r}
p <- ggplot(data = gapminder, mapping = aes(x = lifeExp)) +
  geom_histogram(binwidth = 5)
```

### Histogram
```{r, echo=FALSE}
p
```


### Bar charts

```{r}
p <- ggplot(data = gapminder, mapping = aes(x = continent)) +
  geom_bar()
```

### Bar charts

```{r, echo=FALSE}
p
````


### Combining geoms

We can layer multiple geoms on top of each other:

```{r}
p <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point() + geom_smooth()
```

### Combining geoms
```{r, echo=FALSE}
p
```

### You try!

Using the `mpg` dataset, make a boxplot with the kind of transmission
on the x-axis and the highway mpg on the y

### You try (answers)

```{r}
p <- ggplot(data = mpg, mapping = aes(x = trans, y = hwy)) +
  geom_boxplot()
```

### You try (answers)

```{r}
p
```

### Bar charts (extended)

Fill vs color

What do you expect this to do:

```{r, eval = FALSE}
thirty_million <- 30000000
ggplot(data = gapminder, mapping = aes(x = pop > thirty_million, color = continent)) +
  geom_bar()
```

```{r, include=FALSE}
thirty_million <- 30000000
```

### Use fill for bar charts

```{r, FALSE}
p <- ggplot(data = gapminder, mapping = aes(x = pop > thirty_million,
                                            fill = continent)) +
  geom_bar()
```

### Fill
```{r, echo=FALSE}
p
```

###

Try to make this plot

```{r, echo = FALSE}
ggplot(data = gapminder, mapping = aes(x = pop > thirty_million, fill = continent)) +
  geom_bar(position = "dodge")
```

### Positions
```{r, eval = FALSE}
ggplot(data = gapminder, mapping = aes(x = pop > thirty_million,
                                       fill = continent)) +
  geom_bar(position = "dodge")
```
There's "stack", "dodge", "identity", "fill" - try them all out!
