---
title: "Dealing with Data with R"
author: "J. Alexander Branham"
date: "Fall 2016"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation:
    slide_level: 3
    incremental: true
    latex_engine: xelatex
    fig_width: 8
    fig_height: 3
    fig_caption: false
classoption: aspectratio=169
---


# Data transformation 
### Data transformation

Next up: data transformation. We'll be working with the `flights` data
frame from the `nycflights13` package, so make sure it's installed
then load it:
```{r, include = FALSE}
library(dplyr)
```

```{r}
## install.packages("nycflights13")
library(dplyr) # for data transformation
library(nycflights13) # example data to work with
```

### The data
```{r}
flights
```

### dplyr verbs
* keep only certain variables -`select`
* reorder - `filter`
* reorder rows - `arrange`
* create new variables - `mutate`
* collapse data into summary statistics - `summarize`

* perform the above actions by groups - `group_by`
* Can also `summarize_all` or `mutate_all`

