---
title: "Data visualization with R"
author: "J. Alexander Branham"
date: "Fall 2016"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation:
    slide_level: 3
    incremental: true
    latex_engine: xelatex
    fig_width: 8
    fig_height: 3
    fig_caption: false
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align='center')
```

# Data visualization 
## Data visualization
### Data visualization 
* Plots are (usually) easier to interpret than tables and numbers
* We'll use `ggplot2` to visualize data in R. Make sure you have it
  installed (only the first time) and loaded (every time):

```{r, eval = FALSE}
install.packages("ggplot2")
```

```{r, results = "hide"}
library(ggplot2)
```

### Data visualization 
We'll be working with the `diamonds` dataset, provided by `ggplot2`.
Check out its documentation with `?diamonds`:

1. How many diamonds are in the dataset?
2. What is the name of the variable for the weight of the diamond?
3. What is the code for the worst color of diamond?

### Scatterplots

```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price))
```

# Aesthetics
## Aesthetics
### Color 
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, color = cut))
```

### Shape
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, shape = cut))
```

### Size
```{r, warning = FALSE}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, size = cut))
```


### Alpha
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price, alpha= cut))
```

### You try!
Use the `mpg` dataset from `ggplot2` to:

1. Make a scatterplot with displ on the x-axis and hwy on the y-axis
2. Add color, size, and shape aesthetics 

# Facets
### Facets
Can make facets by adding `facet_grid`:
```{r}
ggplot(data = diamonds) +
  geom_point(mapping = aes(x = carat, y = price)) + facet_grid(. ~ cut)
```

# Geoms
### Geoms
There are other kinds of graphs than scatterplots. `geom_` takes care
of this for ggplot2 ("geometric object"):

### Smooth
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price, color = cut)) +
  geom_smooth()
```

### Combining geoms 
We can layer multiple geoms on top of each other:

```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price)) +
  geom_point() + geom_smooth()
```

### Smooth - SE's
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price, color = cut)) +
  geom_smooth(se = FALSE) 
```

### Smooth - OLS
```{r}
ggplot(data = diamonds, mapping = aes(x = carat, y = price, color = cut)) +
  geom_smooth(se = FALSE, method = "lm") 
```

### Histogram

```{r}
ggplot(data = diamonds, mapping = aes(x = carat)) +
  geom_histogram(binwidth = .25)
```

### Bar charts
```{r}
ggplot(data = diamonds, mapping = aes(x = cut)) +
  geom_bar()
```

### You try!
Make a boxplot with color on the x-axis and price on the y-axis

### You try (answers)

```{r}
ggplot(data = diamonds, mapping = aes(x = color, y = price)) +
  geom_boxplot()
```

### Bar charts (extended)
Fill vs color

What do you expect this to do:

```{r, eval = FALSE}
ggplot(data = diamonds, mapping = aes(x = cut, color = cut)) +
  geom_bar()
```

### Use fill for bar charts

```{r, FALSE}
ggplot(data = diamonds, mapping = aes(x = cut, fill = cut)) +
  geom_bar()
```

### 
Try to make this plot

```{r, echo = FALSE}
ggplot(data = diamonds, mapping = aes(x = color, fill = cut)) +
  geom_bar(position = "dodge")
```

### Positions
```{r, eval = FALSE}
ggplot(data = diamonds, mapping = aes(x = color, fill = cut)) +
  geom_bar(position = "dodge")
```
There's "stack", "dodge", "identity", "fill" - try them all out!
