---
title: "Dealing with Data in R"
subtitle: "How to Use dplyr"
author: "J. Alexander Branham"
date: "Fall 2016"
header-includes: 
  - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation:
    slide_level: 3
    incremental: true
    latex_engine: xelatex
    fig_width: 8
    fig_height: 3
    fig_caption: false
classoption: aspectratio=169
fontsize: bigger
---

```{r, echo=FALSE}
source("r-setup.R")
```

# Data transformation 
### Data transformation

Next up: data transformation. We'll be working with the `gapminder` data
frame from the `gapminder` package, so make sure it's installed
then load it:
```{r}
## install.packages(c("gapminder", "dplyr"))
library(dplyr) # for data transformation
library(gapminder) # example data to work with
```

### The data
```{r}
gapminder
```

### dplyr verbs
* keep only certain observations - `filter`
* keep only certain variables -`select`
* reorder rows - `arrange`
* create new variables - `mutate`
* collapse data into summary statistics - `summarize`

\pause 

Perform the above actions by groups - `group_by`

### dplyr verbs
* Usage for all of these is the same
    + Name of the data frame
    + What to do with the data frame
    + Result is always a data frame

### filter
How to get only countries in Africa?
\pause 

```{r}
filter(gapminder, continent == "Africa")
```

### You try! 
Get a data frame of all the countries in Europe in 1997

### You try (answer)
```{r}
filter(gapminder, continent == "Europe", year == 1997)
```

### Comparisons
R supports several logical comparisons:

* Equal `==`
* Not equal `!=`
* Greater than `>` (or equal to `>=`)
* Less than `<` (or equal to `<=`)

### And, or, and %in%
* `filter` automatically joins multiple arguments with `&`
* You can use `|` instead, which means "or"
* Try to get all the countries in Europe or Africa

### and, or, and %in%
```{r, error = TRUE}
filter(gapminder, continent == "Europe" |  "Africa")
```

### and, or, and %in% 

```{r}
filter(gapminder, continent %in% c("Europe", "Africa"))
```

### What's or used for?

```{r}
filter(gapminder, continent == "Asia" | country == "Turkey")
```

### Select
Sometimes you'll want to keep only the columns you're interested in.
`select` lets you do that:

```{r}
select(gapminder, country, year, pop)
```

### Select helper functions
`select` has some helper functions: `starts_with` and `ends_with` are
among the most useful:

```{r}
select(gapminder, starts_with("c"), pop)
```

### Rename
You can use `select` to rename variables, but since it drops
everything that it doesn't return, it oftentimes isn't good at that.
`rename` does what you want it to, though:

```{r}
rename(gapminder, population = pop)
```

### Arrange
```{r}
arrange(gapminder, year)
```

### Mutate
`mutate` allows you to create new variables:

```{r}
mutate(gapminder, gdp = pop * gdpPercap)
```

### Mutate
We can create multiple variables at once:

```{r}
mutate(gapminder,
       gdp = pop * gdpPercap,
       gdp_in_billions = gdp / 1000000)
```

### Summarize
`summarize` (or `summarise` if you prefer) creates summary statistics:

```{r}
summarize(gapminder, mean_life = mean(lifeExp))
```

### group_by
`group_by` allows us to perform operations by groups:

```{r}
by_year <- group_by(gapminder, year)
summarize(by_year, mean_life = mean(lifeExp))
```

### Piping
The pipe operator `%>%` pipes the output of the left side to the first
argument of the right side:

```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(mean_life = mean(lifeExp),
            n = n())
```

### You try! 

* What is the mean life expectancy in Europe in 1997?
* What is the total population of Asia in 1992?
* Create a plot with year along the x-axis and average life expectancy
by continent along the y-axis. 
  
### You try (answers)
```{r}
gapminder %>%
  filter(year == 1997, continent == "Europe") %>%
  summarize(mean_life = mean(lifeExp))
```

### You try (answers)
```{r}
gapminder %>%
  filter(continent == "Asia", year == 1992) %>%
  summarize(total_pop = sum(as.numeric(pop)))
```

### You try (answers)
```{r}
gapminder %>%
  group_by(year, continent) %>%
  summarize(mean_life = mean(lifeExp)) %>%
  ggplot(aes(year, mean_life, color = continent)) +
  geom_point() + geom_smooth()
```

### Summarize all 
We can use `summarize_all` to summarize multiple variables:

```{r, warnings = FALSE}
gapminder %>%
  group_by(year) %>%
  summarize_all(mean)
```

### Summarize if
`summarize_if` allows us to do conditional summaries:

```{r}
gapminder %>%
  group_by(year) %>%
  summarize_if(is.numeric, mean)
```   
